---
title: "Drinking Water Quality Analysis"
author: "Tom Gibbens-Matsuyama"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

### Load libraries
```{r}
library(tidyverse)
library(here)
```

### Load data
```{r}
lab_water <- read_csv(here("data", "lab_results.csv"))

field_water <- read_csv(here("data", "field_results.csv"))
```

### Explore the data
The variables that I am interested in are parameter, result, station_type, sample date. I am particulary interested in the type of variables that are within the parameter column. Just by looking at the data set from the preview, I can see that pH, hardness, dissolved molecules, and much more are involved. However, I want to see exactly what they have in there since there are over 4 million rows. 

### Explore the data
```{r}
unique(lab_water$parameter)
```

Looking at this dataset, the columns I can see that i am interested in are "pH", "Total Hardness", "Dissolved Hardness", "Turbidity", "Dissolved Mercury", "Dissolved Lead", "Dissolved Hardness"

```{r}
unique(field_water$parameter)
```

Looking at this dataset, the columns I can see that I am interested in are "pH", "DissolvedOxygen", "Turbidity", "Algae (Description)" 

```{r}
unique(field_water$fdr_result)
```

